---
title: "Diagnose Breast Cancer"
author: "Scott Ligon and Karen Tafolla"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective
We will use mammogram screenings from 2,392,998 women in order to predict whether the patient is in risk of breast cancer. 

## Information About Data Set

Each mammogram screening includes the following features:
  
  + Menopause
  + Age Group
  + Brest Density
  + Race
  + Hispanic
  + Body Mass Index
  + Age of First Birth
  + Relatives with Breast Cancer
  + Breast Procedure
  + Last Mammogram Result
  + Surgical Menopause
  + Hormone Therapy
  + Invasive breast cancer
  + Cancer
  + Training
  + Count

Data for this study was obtained from the BCSC Data Resource. More information regarding this resource is available at: http://breastscreening.cancer.gov/.

## Gathering and Processing Data
We modified the feature names in order to have more meaninful names rather than acronyms that might be hard to understand. 

```{r}
mammogram = read.csv("risk.csv")
names(mammogram) <- c("menopause","age.group","breast.density","race","hispanic","bmi","age.first.birth","relatives.with.bc","breast.procedure","last.mammogram","surgical.menopause","hormone.therapy","invasive.bc","cancer","training","count")

mammogram$age.group = factor(mammogram$age.group, levels = c(1,2,3,4,5,6,7,8,9,10), labels = c("35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80-84"))

mammogram$race = factor(mammogram$race, levels = c(1,2,3,4,5,9), labels = c("white", "Asian/Pacific Islander","black", "Native American", "other/mixed", "unknown"))

mammogram$bmi = factor(mammogram$bmi, levels = c(1,2,3,4,9), labels = c("10-24.99", "25-29.99", "30-34.99", "35 or more","unknown"))

mammogram$breast.density = factor(mammogram$breast.density, levels = c(1,2,3,4,9), labels = c("fat","scattered", "dense", "extremely dense", "differnt measurement"))

sum(mammogram$race == "unknown")
sum(mammogram$bmi == "unknown")
sum(mammogram$breast.density == "unknown")

```

```{r}
str(mammogram)
```


```{r}
summary(mammogram)
```

```{r}
to explore

mammogram$age.group
mammogram$breast.density
mammogram$race
mammogram$hispanic
mammogram$hormone.therapy

plot(density(mammogram$hispanic))
```

Possible 
## Logistic Regression

## Classification Trees

## Naive Bayes
